---
title: "HW02p"
author: "Rachel Hia"
date: "March 6, 2018"
output: pdf_document
---

Welcome to HW02p where the "p" stands for "practice" meaning you will use R to solve practical problems. This homework is due 11:59 PM Tuesday 3/6/18. 

You should have RStudio installed to edit this file. You will write code in places marked "TO-DO" to complete the problems. Some of this will be a pure programming assignment. Sometimes you will have to also write English.

The tools for the solutions to these problems can be found in the class practice lectures. I want you to use the methods I taught you, not for you to google and come up with whatever works. You won't learn that way.

To "hand in" the homework, you should compile or publish this file into a PDF that includes output of your code. To do so, use the knit menu in RStudio. You will need LaTeX installed on your computer. See the email announcement I sent out about this. Once it's done, push the PDF file to your github class repository by the deadline. You can choose to make this respository private.

For this homework, you will need the `testthat` libray.

```{r}
pacman::p_load(testthat)
```


1. Source the simple dataset from lecture 6p:

```{r}
Xy_simple = data.frame(
 response = factor(c(0, 0, 0, 1, 1, 1)), #nominal
 first_feature = c(1, 1, 2, 3, 3, 4),    #continuous
 second_feature = c(1, 2, 1, 3, 4, 3)    #continuous
)
X_simple_feature_matrix = as.matrix(Xy_simple[, 2 : 3])
y_binary = as.numeric(Xy_simple$response == 1)
```


Try your best to write a general perceptron learning algorithm to the following `Roxygen` spec. For inspiration, see the one I wrote in lecture 6.

```{r}
#' This function implements the "perceptron learning algorithm" of Frank Rosenblatt (1957).
#'
#' @param Xinput      The training data features as an n x (p + 1) matrix where the first column is all 1's.
#' @param y_binary    The training data responses as a vector of length n consisting of only 0's and 1's.
#' @param MAX_ITER    The maximum number of iterations the perceptron algorithm performs. Defaults to 1000.
#' @param w           A vector of length p + 1 specifying the parameter (weight) starting point. Default is 
#'                    \code{NULL} which means the function employs random standard uniform values.
#' @return            The computed final parameter (weight) as a vector of length p + 1
perceptron_learning_algorithm = function(Xinput, y_binary, MAX_ITER = 1000, w = NULL){
  #TO-DO
  if (is.null(w)){
    w = runif(ncol(Xinput)) #intialize a p+1-dim vector with random values  
  }
  for (iter in 1 : MAX_ITER){  
    for (i in 1 : nrow(Xinput)){
      x_i = Xinput[i, ]
      yhat_i = ifelse(x_i %*% w > 0, 1, 0)
      w = w + as.numeric(y_binary[i] - yhat_i) * x_i
    }
  }
  w
}
```

Run the code on the simple dataset above via:

```{r}yy
w_vec_simple_per = perceptron_learning_algorithm(
  cbind(1, Xy_simple$first_feature, Xy_simple$second_feature),
  as.numeric(Xy_simple$response == 1))
w_vec_simple_per
```

Use the ggplot code to plot the data and the perceptron's $g$ function.

```{r}
pacman::p_load(ggplot2)
simple_viz_obj = ggplot(Xy_simple, aes(x = first_feature, y = second_feature, color = response)) + 
  geom_point(size = 5)
simple_perceptron_line = geom_abline(
    intercept = -w_vec_simple_per[1] / w_vec_simple_per[3], 
    slope = -w_vec_simple_per[2] / w_vec_simple_per[3], 
    color = "orange")
simple_viz_obj + simple_perceptron_line
```

Why is this line of separation not "satisfying" to you?
This line is close to our values. This example is linearly separable so our line looks pretty much in the middle of the two different response color dots.  

TO-DO



2. Use the `e1071` package to fit an SVM model to `y_binary` using the predictors found in `X_simple_feature_matrix`. Do not specify the lambda

```{r}
pacman::p_load(e1071)
Xy_simple_feature_matrix = as.matrix(Xy_simple[, 2 : 3])
lambda = 1e-9
n = nrow(Xy_simple_feature_matrix)
svm_model = #TO-DO 
svm(Xy_simple_feature_matrix, Xy_simple$response, kernel = "linear",  scale = FALSE)
```

and then use the following code to visualize the line in purple:

```{r}
w_vec_simple_svm = c(
  svm_model$rho, #the b term
  -t(svm_model$coefs) %*% X_simple_feature_matrix[svm_model$index, ] # the other terms
)
simple_svm_line = geom_abline(
    intercept = -w_vec_simple_svm[1] / w_vec_simple_svm[3], 
    slope = -w_vec_simple_svm[2] / w_vec_simple_svm[3], 
    color = "purple")
simple_viz_obj + simple_perceptron_line + simple_svm_line
```

Is this SVM line a better fit than the perceptron?
yes. The SVM purple line seems to be more fitting than the orange perceptron line. The SVM line is divided more evenly. The distance between each of the points and the purple line is significantly less than the distance between each of the points and the orange line.

TO-DO

3. Now write your own implementation of the linear support vector machine algorithm respecting the following spec making use of the nelder mead `optim` function from lecture 5p. It turns out you do not need to load the package `neldermead` to use this function. You can feel free to define a function within this function if you wish. 

Note there are differences between this spec and the perceptron learning algorithm spec in question \#1. You should figure out a way to respect the `MAX_ITER` argument value. 

```{r}
#' This function implements the hinge-loss +maximum margin  linear support vector machine algorithm of Vladimir Vapnik (1963).
#'
#' @param Xinput      The training data features as an n x p matrix.
#' @param y_binary    The training data responses as a vector of length n consisting of only 0's and 1's.
#' @param MAX_ITER    The maximum number of iterations the algorithm performs. Defaults to 5000.
#' @param lambda      A scalar hyperparameter trading off margin of the hyperplane versus average hinge loss.
#'                    The default value is 0.1 to mimic hard margin in the linearly separable case.
#' @return            The computed final parameter (weight) as a vector of length p + 1
linear_svm_learning_algorithm = function(Xinput, y_binary, MAX_ITER = 5000, lambda = .1){
  #TO-DO
  #minimum avg hinge loss + maximum margin
  #make a cost function and input it in optim
  #if it doesn't work, don't stress about it, there's something wrong with optim
  p = ncol(Xinput)
  Xinput = cbind(1, Xinput)
  w_0 = rep(0, p + 1)
  optim(w_0, function(w){
    average_hinge_loss = 1 / nrow(Xinput) * sum(
      pmax(0, 0.5 - (y_binary - 0.5) * (Xinput %*% w)))
    scaled_inverse_margin_width = lambda * sum(w[2 : (p + 1)]^2)
    average_hinge_loss + scaled_inverse_margin_width
  },
  control = list(maxit = MAX_ITER))$par
 
}
  
#initialize w and run optim
# optim_output = optim(c(0, 0), SAE)
# optim_output
# w_vec = optim_output$par
# }
```

Run your function using the defaults and plot it in brown vis-a-vis the previous model's line:

```{r}
svm_model_weights = linear_svm_learning_algorithm(X_simple_feature_matrix, y_binary)
my_svm_line = geom_abline(
    intercept = -svm_model_weights[1] / svm_model_weights[3], 
    slope = -svm_model_weights[2] / svm_model_weights[3], 
    color = "brown")
simple_viz_obj + simple_svm_line + my_svm_line
```

Is this the same as what the `e1071` implementation returned? Why or why not?
They look very similar. They look to have the same slope. However, the brown and the purple line are not the same. They probably have a different y-intercept. They look to be parallel and close together, but different. 

4. Write a $k=1$ nearest neighbor algorithm using the Euclidean distance function. Respect the spec below:

```{r}
#' This function implements the nearest neighbor algorithm.
#'
#' @param Xinput      The training data features as an n x p matrix.
#' @param y_binary    The training data responses as a vector of length n consisting of only 0's and 1's.
#' @param Xtest       The test data that the algorithm will predict on as a n* x p matrix.
#' @return            The predictions as a n* length vector.
nn_algorithm_predict = function(Xinput, y_binary, Xtest){ #first row, dot with second row
#TO-DO
r=rep(NA,nrow(Xtest))
best_sqd_distance = 1000
i_star = NA 
for (i in 1 : nrow(Xtest) ){ 
  for(index in 1:nrow(Xinput)){ 
    eucl = sqrt(sum((Xinput[i,] - Xtest[i,])^2)) 
    if (eucl < best_sqd_distance){ 
      best_sqd_distance = eucl 
      i_star = i 
      r[i]=y_binary[i_star] }
  } 
  best_sqd_distance= Inf } 
r}
```

Write a few tests to ensure it actually works:

```{r}
#TO-DO
nntest = nn_algorithm_predict(X_simple_feature_matrix, y_binary, X_simple_feature_matrix);
expect_equal(nntest, y_binary)

```

For extra credit, add an argument `k` to the `nn_algorithm_predict` function and update the implementation so it performs KNN. In the case of a tie, choose $\hat{y}$ randomly. Set the default `k` to be the square root of the size of $\mathcal{D}$ which is an empirical rule-of-thumb popularized by the "Pattern Classification" book by Duda, Hart and Stork (2007). Also, alter the documentation in the appropriate places.

```{r}
#not required TO-DO --- only for extra credit
```

For extra credit, in addition to the argument `k`, add an argument `d` representing any legal distance function to the `nn_algorithm_predict` function. Update the implementation so it performs KNN using that distance function. Set the default function to be the Euclidean distance in the original function. Also, alter the documentation in the appropriate places.

```{r}
#not required TO-DO --- only for extra credit
```


5. We move on to simple linear modeling using the ordinary least squares algorithm.

Let's quickly recreate the sample data set from practice lecture 7:

```{r}
n = 20
x = runif(n)
beta_0 = 3
beta_1 = -2
y = beta_0 + beta_1 * x + rnorm(n, mean = 0, sd = 0.33)
```

Solve for the least squares line by computing $b_0$ and $b_1$ _without_ using the functions `cor`, `cov`, `var`, `sd` but instead computing it from the $x$ and $y$ quantities manually. See the class notes.

```{r}
#TO-DO
ybar = mean(y)
xbar = mean(x)
s_x = sqrt(sum((x-mean(x))^2/(length(x)-1)))
s_y = sqrt(sum((y-mean(y))^2/(length(y)-1)))
xy = x*y
s_xy = (sum(x*y) - (length(x) *xbar *ybar)) / (length(x)-1)

r = s_xy / (s_x * s_y)

b_1 = r * s_y / s_x
b_0 = ybar - b_1 * xbar
b_0
b_1
```

Verify your computations are correct using the `lm` function in R:

```{r}
pacman::p_load(testthat)
lm_mod = lm(y ~ x)
b_vec = coef(lm_mod)
b_vec
#this test doesnt work yet
expect_equal(b_0, as.numeric(b_vec[1]), tol = 1e-4)
expect_equal(b_1, as.numeric(b_vec[2]), tol = 1e-4)
```

6. We are now going to repeat one of the first linear model building exercises in history --- that of Sir Francis Galton in 1886. First load up package `HistData`.

```{r}
#TO-DO
if(!require("HistData"))(install.packages("HistData"))

```

In it, there is a dataset called `Galton`. Load it using the `data` command:

```{r}
#TO-DO
data(Galton)
```

You now should have a data frame in your workspace called `Galton`. Summarize this data frame and write a few sentences about what you see. Make sure you report $n$, $p$ and a bit about what the columns represent and how the data was measured. See the help file `?Galton`.

```{r}
#TO-DO
summary(Galton)
View(Galton)

```

TO-DO
This dataset has 2 columns. One for height of the parents and one for the height of the child. I think this dataset could be used to predict a child's height given the parent's height. The heights are measured in inches. The height of the parent is the average height of the father and mother. The heights of the parents and children here are very close because the means are very similar. The dataset includes A data frame with 928 observations on 2 variables. p is 1 and n is 928.

Find the average height (include both parents and children in this computation).

```{r}
avg_height = (sum(Galton$parent + Galton$child))/ (length(Galton$parent) +length(Galton$child))
avg_height
  
```

Note that in Math 241 you learned that the sample average is an estimate of the "mean", the population expected value of height. We will call the average the "mean" going forward since it is probably correct to the nearest tenth of an inch with this amount of data.

Run a linear model attempting to explain the childrens' height using the parents' height. Use `lm` and use the R formula notation. Compute and report $b_0$, $b_1$, RMSE and $R^2$. Use the correct units to report these quantities.

```{r}
#TO-DO
mod = lm(Galton$parent ~ Galton$child, data = Galton)
coef(mod)
summary(mod)$sigma #RMSE
summary(mod)$r.squared
```

Interpret all four quantities: $b_0$, $b_1$, RMSE and $R^2$.

TO-DO
$b_0$ is our intercept of our prediction line. 
$b_1$ is the slope of our prediction line.
RMSE is 1.58. This means that if we're using logic similar to a confidence integral, 95 percent of our predictions will be about 3 inches from what we predict. 
$R^2$ is 0.21. This data tells us how close the data is to the prediction line. It is not as useful as RMSE is in practical applications.


How good is this model? How well does it predict? Discuss.
This model doesn't seem very accurate. Our RMSE tells us that our predictions will be about 3 inches from what we predict. That seems like a lot considering the differences between heights are not that big. 

TO-DO



Now use the code from practice lecture 8 to plot the data and a best fit line using package `ggplot2`. Don't forget to load the library.

```{r}
#TO-DO
pacman::p_load(ggplot2)
ggplot(Galton, aes(parent, child)) + 
  geom_point() + 
  geom_smooth(method = 'lm')
b= coef(mod)
```

It is reasonable to assume that parents and their children have the same height. Explain why this is reasonable using basic biology.
Children inherit their DNA from their parents. If the parents are both short, thwn the child is likely to be short. If the parents are both tall, then the child is likely to be tall. Although genetics is a common predictor, there are many other factors that could predict human development such as nutrition and exercise and lack of certain vitamins and minerals. Just using the parent's height alone may not be the best predictor.  

TO-DO

If they were to have the same height and any differences were just random noise with expectation 0, what would the values of $\beta_0$ and $\beta_1$ be?

beta_0 is the intercept of the prediction line. In this case, the predictions would be reality because there are no differences between parent's height and child's height. 
beta_1 is the slope of our prediction line. This slope would be 1. 


TO-DO

Let's plot (a) the data in $\mathbb{D}$ as black dots, (b) your least squares line defined by $b_0$ and $b_1$ in blue, (c) the theoretical line $\beta_0$ and $\beta_1$ if the parent-child height equality held in red and (d) the mean height in green.

```{r}
ggplot(Galton, aes(x = parent, y = child)) + 
  geom_point() + 
  geom_jitter() +
  geom_abline(intercept = b[1], slope = b[2], color = "blue", size = 1) +
  geom_abline(intercept = 0, slope = 1, color = "red", size = 1) +
  geom_abline(intercept = avg_height, slope = 0, color = "darkgreen", size = 1) +
  xlim(63.5, 72.5) + 
  ylim(63.5, 72.5) +
  coord_equal(ratio = 1)
```

Fill in the following sentence: 

TO-DO: Children of short parents became short on average and children of tall parents became tall on average.

Why did Galton call it "Regression towards mediocrity in hereditary stature" which was later shortened to "regression to the mean"?

TO-DO
Mediocrity is average or mean. Hereditary stature is just how tall someone is. Regression towards mediocrity means a measure of the relation between the mean value of one varaible and corresponding values of other variables. The regression is unstable in the short run and after a while, it will close to the mean.

Why should this effect be real?

TO-DO
Any single measurement could be an outliar and not even close to the mean. Once we have a large enough n, our regression line gets closer to the mean. This effect is real because our data won't be true of any one child's height but could give us an estimate about the average
child's height.



You now have unlocked the mystery. Why is it that when modeling with $y$ continuous, everyone calls it "regression"? Write a better, more descriptive and appropriate name for building predictive models with $y$ continuous.

TO-DO
Regression is a statistical process for estimating relationships between variables. Regression implies the data will behave in a certain way, ie mediocrity. This is an estimation. We could have a lot of really tall people and a lot of really short people with little average height people and our data will predict most people to be average. This is an indication that the term 'regression' will simplify our data into something, in this case a line, that can help us estimate the next child's height. I would call it an estimation tool. 

